---
title: "Lab 10"
subtitle: Due Thursday April 9 at 11:59 PM
format: html
draft: false
editor: 
  mode: source
---

This lab provides extra practice with the kind of maximum likelihood problem that you will encounter on the last problem set and the final exam.

# Task 0 (the TA does it)

Let's see a worked example first. Imagine we observe data 

$$
X_1,\,X_2,\,...,\,X_n\overset{\text{iid}}{\sim}\text{N}(\theta, 1),
$$

where the true value of the mean $\theta>0$ is unknown. We will use the method of maximum likelihood to propose an estimator for $\theta$. Spoiler alert: In this case, it's the sample mean $\bar{X}_{n}$! See below for the derivation.

::: {.callout-warning collapse="true"}
## Derivation

We consider data 

$$
X_1,\,X_2,\,...,\,X_n\overset{\text{iid}}{\sim}\text{N}(\theta, 1).
$$
Recall that the associated density is given by $$f(x;\ \theta) = \frac{1}{\sqrt{2\pi}}e^{-\frac{(x - \theta)^{2}}{2}}, \quad x \in \mathbb{R}.$$ So, the joint likelihood function is given by
$$
\begin{aligned}
L\left(\theta;\ X_{1:n}\right) & = \prod_{i = 1}^{n}{f\left(X_{i};\ \theta\right)} \\ & = \prod_{i = 1}^{n}{\frac{1}{\sqrt{2\pi}}e^{-\frac{\left(X_{i} - \theta\right)^{2}}{2}}} \\ & = (2\pi)^{-\frac{n}{2}}\prod_{i = 1}^{n}{e^{-\frac{\left(X_{i} - \theta\right)^{2}}{2}}} \\ & = (2\pi)^{-\frac{n}{2}}e^{-\frac{1}{2}\sum_{i = 1}^{n}{\left(X_{i} - \theta\right)^{2}}}.
\end{aligned}
$$
The log-likelihood is then given by
$$
\begin{aligned}
\ell\left(\theta;\ X_{1:n}\right) & = \ln L\left(\theta; \ X_{1:n}\right) \\ & = \ln\left[(2\pi)^{-\frac{n}{2}}e^{-\frac{1}{2}\sum_{i = 1}^{n}{\left(X_{i} - \theta\right)^{2}}}\right] \\ & = \ln\left[(2\pi)^{-\frac{n}{2}}\right] + \ln\left[e^{-\frac{1}{2}\sum_{i = 1}^{n}{\left(X_{i} - \theta\right)^{2}}}\right] \\ & = -\frac{n}{2}\ln(2\pi) - \frac{1}{2}\sum_{i = 1}^{n}{\left(X_{i} - \theta\right)^{2}}.
\end{aligned}
$$
We want to find $$\hat{\theta}_{n} = \underset{\theta}{\arg\max}{\ L\left(\theta;\ X_{1:n}\right)} = \underset{\theta}{\arg\max}{\ \ell\left(\theta;\ X_{1:n}\right)}.$$ We optimize by taking the derivative with respect to $\theta$, setting it equal to zero, and solving:
$$
\begin{aligned}
\frac{\text{d}\ell}{\text{d}\theta} & = \frac{\text{d}}{\text{d}\theta}\left[-\frac{n}{2}\ln(2\pi) - \frac{1}{2}\sum_{i = 1}^{n}{\left(X_{i} - \theta\right)^{2}}\right] \\ & = -\frac{1}{2}\sum_{i = 1}^{n}{\frac{\text{d}}{\text{d}\theta}\left[\left(X_{i} - \theta\right)^{2}\right]} \\ & = -\frac{1}{2}\sum_{i = 1}^{n}{-2\left(X_{i} - \theta\right)} \\ & = \sum_{i = 1}^{n}{X_{i}} - n\theta \\ \Rightarrow 0 & = \sum_{i = 1}^{n}{X_{i}} - n\hat{\theta}_{n} \\ \Rightarrow n\hat{\theta}_{n} & = \sum_{i = 1}^{n}{X_{i}} \\ \Rightarrow \hat{\theta}_{n} & = \frac{1}{n}\sum_{i = 1}^{n}{X_{i}} \\ \Rightarrow \hat{\theta}_{n} & = \bar{X}_{n}.
\end{aligned}
$$
For the sake of completeness, we can check the second derivative to verify that this is, in fact, a maximum:
$$
\begin{aligned}
\frac{\text{d}^{2}\ell}{\text{d}\theta^{2}} & = \frac{\text{d}}{\text{d}\theta}\left[\frac{\text{d}\ell}{\text{d}\theta}\right] \\ & = \frac{\text{d}}{\text{d}\theta}\left[\sum_{i = 1}^{n}{X_{i}} - n\theta\right] \\ & = -n.
\end{aligned}
$$
Since $\frac{\text{d}^{2}\ell}{\text{d}\theta^{2}} = -n < 0$, we know that $\hat{\theta}_{n} = \bar{X}_{n}$ is the *maximum* likelihood estimator. 
:::

# Task 1 (you do it)

We now pivot to our distribution of choice for this lab. If a non-negative random variable $X$ has the *Rayleigh distribution* with parameter $\theta>0$, then its CDF is 

$$
F(x;\,\theta)=1-\exp\left(-\frac{x^2}{2\theta}\right)\quad x\geq 0,
$$

and we say $X\sim\text{Rayleigh}(\theta)$. What is the PDF in this parametric family?

<!--
::: {.callout-warning collapse="true"}
## Solution

$$
\begin{aligned}
f(x;\,\theta)
&=
F'(x;\,\theta)
\\
&=
\frac{\text{d}}{\text{d}x}
\left[
1-\exp\left(-\frac{x^2}{2\theta}\right)
\right]
\\
&=
0-\frac{-x}{\theta}
\exp\left(-\frac{x^2}{2\theta}\right)
\\
&=
\frac{x}{\theta}
\exp\left(-\frac{x^2}{2\theta}\right),&&x\geq 0.
\end{aligned}
$$

:::
-->

# Task 2 (we do it)

Imagine we observe data 

$$
X_1,\,X_2,\,...,\,X_n\overset{\text{iid}}{\sim}\text{Rayleigh}(\theta),
$$

where the true value of the parameter $\theta>0$ is unknown. Use the method of maximum likelihood to propose an estimator for $\theta$.

::: {.callout-warning collapse="true"}
## Solution
The likelihood function is 

$$
\begin{aligned}
L(\theta;\,X_{1:n})
&=
\prod_{i=1}^n
f(X_i;\,\theta)
\\
&=
\prod_{i=1}^n
\frac{X_i}{\theta}
\exp\left(-\frac{X_i^2}{2\theta}\right)
\\
&=
\theta^{-n}\left(\prod_{i=1}^nX_i\right)\left[\prod_{i=1}^n\exp\left(-\frac{X_i^2}{2\theta}\right)\right]
\\
&=
\theta^{-n}\left(\prod_{i=1}^nX_i\right)\exp\left(-\frac{1}{2\theta}\sum\limits_{i=1}^nX_i^2\right).
\end{aligned}
$$

The log-likelihood function is 

$$
\begin{aligned}
\ell(\theta;\,X_{1:n})
&=
\ln L(\theta;\,X_{1:n})
\\
&=
\ln\left[\theta^{-n}\left(\prod_{i=1}^nX_i\right)\exp\left(-\frac{1}{2\theta}\sum\limits_{i=1}^nX_i^2\right)\right]
\\
&=
\ln(\theta^{-n})+\ln\left(\prod_{i=1}^nX_i\right)+\ln\left[\exp\left(-\frac{1}{2\theta}\sum\limits_{i=1}^nX_i^2\right)\right]
\\
&=
-n\ln \theta+\sum\limits_{i=1}^n\ln X_i-\frac{1}{2\theta}\sum\limits_{i=1}^nX_i^2
.
\end{aligned}
$$

To maximize this, we take the second derivative, set it equal to zero, and solve:

$$
\begin{aligned}
\frac{\text{d}}{\text{d}\theta}\ell(\theta;\,X_{1:n})
=
-
\frac{n}{\theta}
+
\frac{1}{2\theta^2}
\sum\limits_{i=1}^nX_i^2
&=
0
\\
\frac{1}{2\theta^2}
\sum\limits_{i=1}^nX_i^2
&=
\frac{n}{\theta}
\\
\frac{1}{2n}
\sum\limits_{i=1}^nX_i^2
&=
\theta.
\end{aligned}
$$

So the maximum likelihood estimator for the Rayleigh distribution is 

$$
\hat{\theta}_n=\frac{1}{2n}
\sum\limits_{i=1}^nX_i^2.
$$

:::

# Task 3 (we do it)

Your estimator in the previous part is a random variable that inherits its randomness from the $X_i$. Derive the full *sampling distribution* of your estimator.

::: {.callout-warning collapse="true"}
## Solution
For these problems, I work "from the inside out." What's the distribution of $X_i$? What's the distribution of $X_i^2$? What's the distribution when you add those up? What's the distribution when you scale the sum by $1/2n$? To answer those questions, you need tools like the [change-of-variables formulas](https://anyakatsevich.github.io/sta240-s26.github.io/labs/lab-7.html) and our [MGF updating rules for linear combinations](https://anyakatsevich.github.io/sta240-s26.github.io/lecture-notes/sta240-notes-14-sums-and-averages.pdf).

We already know that $X_1\sim\text{Rayleigh}(\theta)$, so we derive the distribution of $Y_1=g(X_1)=X_1^2$ using change of variables. Since $\text{Range}(X_1)=(0,\,\infty)$, the transformation is strictly increasing, so we apply the change of variables formula:

$$
\begin{aligned}
f_{Y_1}(y)
&=
f\left(g^{-1}(y);\,\theta\right)\left|\frac{\text{d}}{\text{d}y}g^{-1}(y)\right|
\\
&=
f\left(y^{1/2};\,\theta\right)\left|\frac{\text{d}}{\text{d}y}y^{1/2}\right|
\\
&=
\frac{y^{1/2}}{\theta}
\exp\left(-\frac{(y^{1/2})^2}{2\theta}\right)
\left|\frac{1}{2}y^{-1/2}\right|
\\
&=
\frac{y^{1/2}y^{-1/2}}{2\theta}
\exp\left(-\frac{y}{2\theta}\right)
\\
&=
\frac{1}{2\theta}
\exp\left(-\frac{y}{2\theta}\right)
,&&y>0.
\end{aligned}
$$

This is the density of $\text{Exponential}(1/2\theta)$, otherwise known as $\text{Gamma}(1,\,1/2\theta)$. Given this, we have the following:

$$
\begin{aligned}
X_i\overset{\text{iid}}{\sim}\text{Rayleigh}(\theta) 
\quad
&\implies
\quad 
X_i^2\overset{\text{iid}}{\sim}\text{Gamma}(1,\,1/2\theta)
\\
&\implies 
\sum\limits_{i=1}^nX_i^2\sim\text{Gamma}(n,\,1/2\theta)
\\
&\implies 
\frac{1}{2n}\sum\limits_{i=1}^nX_i^2\sim\text{Gamma}(n,\,n/\theta).
\end{aligned}
$$
The last implication follows from [Problem Set 5 #10d](https://anyakatsevich.github.io/sta240-s26.github.io/problems/pset-5.html#problem-10). So, the full sampling distribution of the estimator is 

$$
\hat{\theta}_n\sim \text{Gamma}(n,\,n/\theta).
$$

:::

# Task 4 (we do it)

Now that you know the sampling distribution of the estimator, compute its mean squared error (MSE) using the bias-variance decomposition. Is the estimator unbiased? Is it consistent?

::: {.callout-warning collapse="true"}
## Solution
Using the mean and variance formulas for the gamma distribution, we see that 

$$
\begin{aligned}
E(\hat{\theta}_n)&=\frac{n}{n/\theta}=\theta\\
\text{var}(\hat{\theta}_n)&=\frac{n}{(n/\theta)^2}=\frac{\theta^2}{n}.
\end{aligned}
$$

Since $E(\hat{\theta}_n)=\theta$, the bias of the estimator is zero, and so the mean-squared-error is simply $\text{var}(\hat{\theta}_n)=\theta^2/n$. As $n\to\infty$, this goes to zero. As such, the estimator is unbiased and consistent.
:::

# Task 5 (you do it)

Compute the quantile function of the Rayleigh distribution and use it to generate 100 draws from the distribution that has $\theta=4$.

<!--
::: {.callout-warning collapse="true"}
## Solution
The quantile function is the inverse cdf, so we solve:

$$
\begin{aligned}
y
&=
1-\exp\left(-\frac{x^2}{2\theta}\right)
\\
\exp\left(-\frac{x^2}{2\theta}\right)
&=
1-y
\\
-\frac{x^2}{2\theta}
&=
\ln(1-y)
\\
x^2
&=
-2\theta\ln(1-y)
\\
x
&=
\sqrt{-2\theta\ln(1-y)}
.
\end{aligned}
$$

So $F^{-1}(y;\,\theta)=\sqrt{-2\theta\ln(1-y)}$. As such, we can implement an inverse transform sampler to simulate from the distribution:

```{r}
set.seed(2345)
n <- 100
theta <- 4
X <- sqrt(-2 * theta * log(1 - runif(n)))
hist(X, breaks = "Scott", freq = FALSE)
curve(exp(-x^2/(2*theta)) * x / theta, 
      from = 0, to = max(X), n = 1000, col = "red", add = TRUE)
```
:::
-->

# Task 6 (you do it)

Write a for loop that simulates the sampling distribution for your estimator. Let $\theta_{0} = 4$. Here is a schematic of what you are doing:

$$
\begin{matrix}
\text{0. True parameter value} &&& \theta_0 && \\
&&& \downarrow && \\
\text{1. True distribution} &&& \text{Rayleigh}(\theta_0) && \\
&\swarrow &\swarrow& \cdots &\searrow&\searrow \\
\text{3. Simulated data}&x_{1:n}^{(1)} &x_{1:n}^{(2)}& \cdots &x_{1:n}^{(M-1)}&x_{1:n}^{(M)} \\
&\downarrow &\downarrow& \cdots &\downarrow&\downarrow \\
\text{4. Different estimates}&\hat{\theta}_n^{(1)} &\hat{\theta}_n^{(2)}& \cdots &\hat{\theta}_n^{(M-1)}&\hat{\theta}_n^{(M)} \\
&\searrow &\searrow& \cdots &\swarrow&\swarrow \\
& && \text{Histogram} && \\
\end{matrix}
$$

At the end, add on top of the histogram a curve of the density that you derived in Task 3. They ought to match.

::: callout-tip
## Template

Here is some template code you can fill in:

```{r}
#| eval: false
set.seed(123)
true_theta <- 4      # true value of parameter
n <- 25                 # sample size
M <- 2500               # how many repetitions
estimates <- numeric(M) # preallocate storage for the sample of estimates

for(m in 1:M){
  # Step 1: simulate a new fake dataset of size n
  # Step 2: apply your formula from Task 2 to compute new estimate
  # Step 3: store it
}

# Step 4: plot histogram of estimates
# Step 5: add density curve of exact sampling distribution

```

:::

<!--
::: {.callout-warning collapse="true"}
## Solution
```{r}
set.seed(123)
true_theta <- 4      # true value of parameter
n <- 25                 # sample size
M <- 2500               # how many repetitions
estimates <- numeric(M) # preallocate storage for the sample of estimates

for(m in 1:M){
  X <- sqrt(-2 * true_theta * log(1 - runif(n)))
  estimates[m] <- sum(X^2) / (2*n)
}

hist(estimates, breaks = "Scott", freq = FALSE)
curve(dgamma(x, shape = n, rate = n / true_theta), 
      from = 0, to = max(estimates), n = 1000, col = "red", 
      lwd = 2, add = TRUE)
```
:::
-->